Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-14 Manuscript/sec2-review-of-graph-neural-network.tex

\begin{table}
    \begin{footnotesize}
        \begin{tabular}{cp{20em}r}
		\toprule
		GNN &
		$\gamma$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		$\boldsymbol{z}_i^{(k)} = 2\boldsymbol{s}^{(k)}_{i} - \boldsymbol{z}_i^{(k-2)}$&
		$O(d_{out})$ \\
		\textbf{GCN} \cite{kipf2017_gcn} &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		AGCN &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE &	
		$\boldsymbol{h}_i^{l+1} =   \delta(\textcolor{blue}{\boldsymbol{W}}^l  [\boldsymbol{s}_i^{l} \parallel \boldsymbol{h}_i^l])$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE-pool &	
		$\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(1)$ \\
		Neural FPs &
		$\boldsymbol{h}_i^{l+1} = \delta(\textcolor{blue}{\boldsymbol{W}}^{l, |\mathcal{N}(i)|}  (\boldsymbol{h}_i^l + \boldsymbol{s}_i^{l}))$ &
		$O(d_{in} * d_{out})$ \\
		SSE &
		$\boldsymbol{h}_i^{l+1} = (1 - \alpha)  \boldsymbol{h}_i^l +\alpha    \delta(\textcolor{blue}{\boldsymbol{W}}^l_1 \delta(\textcolor{blue}{\boldsymbol{W}}^l_2 [\boldsymbol{x}_i \parallel \boldsymbol{s}_i^l]))$ &
		$O((f + d_{in}) * d_{out})$ \\
		\textbf{GGNN} &
		\begin{scriptsize}
		$\begin{aligned}[t]
			& {\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz}) \\ 
			&\boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr}) \\ 
			&\boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}} \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l}) + \textcolor{blue}{\boldsymbol{b}}^h)) \\ 
			&\boldsymbol{h}_i^{l+1} = (1 - {\boldsymbol{z}}_i^l) \odot \boldsymbol{h}_i^l + {\boldsymbol{z}}_i^l \odot \boldsymbol{h}_i^{l+1}
		\end{aligned}$
		\end{scriptsize}
		&
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		$\hat{\boldsymbol{h}}^{l,k}_{i}} = \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_i^{l}$ (pre-edge), $\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\boldsymbol{g}_i = \textcolor{blue}{\boldsymbol{W}}^l_g  [\boldsymbol{h}_i^{l} \parallel \boldsymbol{s}_{i, 2}^l \parallel \boldsymbol{s}_{i, 3}^l]  \\
			&\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l_o [\boldsymbol{h}_i^l \parallel (\boldsymbol{g}_{i} \odot \boldsymbol{s}_{i, 1}^l) ]
			\end{aligned}$ 
		\end{scriptsize}&
		$O(max(K * d_v + d_{in}, 2 * d_{in} + d_m) * d_{out})$ \\ \bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their vertex computation functions. $d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.For Neural FPs, $\textcolor{blue}{\boldsymbol{W}}^{l, |\mathcal{N}(i)|}$ is the weight matrix for nodes with degree
	$|\mathcal{N}(i)|$ at layer $l$.}
	\label{tab:gnn_overview_vertex}
\end{table}

@@

\textbf{GAT} (low vertex \& high edge computational complexity):
GAT introduces the attention mechanism into the graph neural networks.
During the message passing, it dynamically determines the weight for an edge based on the hidden feature vectors of its two end vertices.
Moreover, it adopts the multi-head mechanism that generates $K$ independent views for an edge, where $K$ is a hyperparameter.
In its message function $\phi$.

>>>>

\begin{table}
    \begin{footnotesize}
        \begin{tabular}{cp{20em}r}
		\toprule
		GNN &
		$\gamma$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		$\boldsymbol{z}_i^{(k)} = 2\boldsymbol{s}^{(k)}_{i} - \boldsymbol{z}_i^{(k-2)}$&
		$O(d_{out})$ \\
		\textbf{GCN} \cite{kipf2017_gcn} &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		AGCN &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE &	
		$\boldsymbol{h}_i^{l+1} =   \delta(\textcolor{blue}{\boldsymbol{W}}^l  [\boldsymbol{s}_i^{l} \parallel \boldsymbol{h}_i^l])$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE-pool &	
		$\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(1)$ \\
		Neural FPs &
		$\boldsymbol{h}_i^{l+1} = \delta(\textcolor{blue}{\boldsymbol{W}}^{l, |\mathcal{N}(i)|}  (\boldsymbol{h}_i^l + \boldsymbol{s}_i^{l}))$ &
		$O(d_{in} * d_{out})$ \\
		SSE &
		$\boldsymbol{h}_i^{l+1} = (1 - \alpha)  \boldsymbol{h}_i^l +\alpha    \delta(\textcolor{blue}{\boldsymbol{W}}^l_1 \delta(\textcolor{blue}{\boldsymbol{W}}^l_2 [\boldsymbol{x}_i \parallel \boldsymbol{s}_i^l]))$ &
		$O((f + d_{in}) * d_{out})$ \\
		\textbf{GGNN} &
		\begin{scriptsize}
		$\begin{aligned}[t]
			& {\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz}) \\ 
			&\boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr}) \\ 
			&\boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}} \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l}) + \textcolor{blue}{\boldsymbol{b}}^h)) \\ 
			&\boldsymbol{h}_i^{l+1} = (1 - {\boldsymbol{z}}_i^l) \odot \boldsymbol{h}_i^l + {\boldsymbol{z}}_i^l \odot \boldsymbol{h}_i^{l+1}
		\end{aligned}$
		\end{scriptsize}
		&
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		$\hat{\boldsymbol{h}}^{l,k}_{i} = \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_i^{l}$ (pre-edge), $\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\boldsymbol{g}_i = \textcolor{blue}{\boldsymbol{W}}^l_g  [\boldsymbol{h}_i^{l} \parallel \boldsymbol{s}_{i, 2}^l \parallel \boldsymbol{s}_{i, 3}^l]  \\
			&\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l_o [\boldsymbol{h}_i^l \parallel (\boldsymbol{g}_{i} \odot \boldsymbol{s}_{i, 1}^l) ]
			\end{aligned}$ 
		\end{scriptsize}&
		$O(max(K * d_v + d_{in}, 2 * d_{in} + d_m) * d_{out})$ \\ \bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their vertex computation functions. $d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.For Neural FPs, $\textcolor{blue}{\boldsymbol{W}}^{l, |\mathcal{N}(i)|}$ is the weight matrix for nodes with degree
	$|\mathcal{N}(i)|$ at layer $l$.}
	\label{tab:gnn_overview_vertex}
\end{table}

@@

\textbf{GAT} (low vertex \& high edge computational complexity):
GAT introduces the attention mechanism into the graph neural networks.
During the message passing, it dynamically determines the weight for an edge based on the hidden feature vectors of its two end vertices.
Moreover, it adopts the multi-head mechanism that generates $K$ independent views for an edge, where $K$ is a hyperparameter.
GAT moves the vertex update function $\gamma$ before the edge computation. 
$\hat{\boldsymbol{h}}^{l,k}_{i} = \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_i^{l}$ with $1 \leq k \leq K$ are conducted for every vertex $v_i$ before the message function $\phi$.
$\phi$ uses $\hat{\boldsymbol{h}}^{l,k}_{i}$ to get the weight of each edge in every head $k$.
After the edge computation, GAT uses the aggreated vector $\boldsymbol{s}^l_i$ directly as the output hidden vector.

\textbf{GaAN} (high vertex \& edge computational complexity):
Based on the multi-head machanism, GaAN introduces a convolutional subnetwork to control the weight of each head.
Thus, it introduces more edge-related parameters to train.
Moreover, GaAN aggregates three messages simutaneously, further increasing the edge computational complexity.

\subsection{Sampling Techniques}

By default, GNNs are trained in a full-batch way, using the whole graph in each iteration.
The full-batch gradient descent has two disadvantages \cite{chiang2019_cluster_gcn}.
It has to cache intermediate results of all vertices in the forward phase, which consumes lots of memory space.
It updates the parameters only once for each epoch, slowing the convergence of gradient descent.

To train GNNs in a mini-batch way, several sampling techniques \cite{chiang2019_cluster_gcn} are proposed.
In each mini-batch, they sample a small subgraph from the whole graph $\mathcal{G}$ and uses the subgraph to update the model parameters.
In other words, the sampling techniques only active the graph neurons and the connections that appear in the sampled subgraph between GNN layers, as shown in \figurename~\ref{fig:graph_sampling}.
Inactive graph neurons and connetions do not participate in the training of this mini-batch, saving lots of computation and storage costs.
Moreover, it may reduce the risk of overfitting the training graph.
Based on whether different GNN layers sample different subgraphs, the existing sampling techniques can be devided into two groups \cite{zeng2020_graphsaint}: the neighbor sampling and the graph sampling.

The neighbor sampling techniques \cite{hamilton2017_graphsage, ying2018_pinsage, chen2018_fastgcn, chen2018_sgcn, huang2018_adap} sample the subgraphs layer by layer.
They first sample several vertices from $\mathcal{V}$ in the last GNN layer.
Then they repeatedly sample the neighbors of those vertices in the previous layer until the input layer.
The sampled subgraphs of different GNN layers may be \emph{different}, as shown in \figurename~\ref{fig:gnn_sampling_neigbhor_sampling}.
GraphSAGE \cite{hamilton2017_graphsage} is the representative technique.
For every sampled vertex $v_i$ in the GNN layer $l$, GraphSAGE samples at most $S^l$ neighbors of $v_i$ in the previous GNN layer. 
$S^l$ is the hyperparameter that is usually much smaller than $|\mathcal{V}|$.
By this way, GraphSAGE limits the neighborhood sizes of the vertices in the sampled subgraph, especially high-degree vertices.

\begin{figure}
	\centering
	\subfloat[Neighbor sampling\label{fig:gnn_sampling_neigbhor_sampling}]{\includegraphics[height=4cm]{figs/illustration/layer_sampling.png}}
	%
	\subfloat[Graph sampling\label{fig:gnn_sampling_graph_sampling}]{\includegraphics[height=4cm]{figs/illustration/graph_sampling.png}}
	\caption{Training a GNN with sampling techniques. The faded graph neurons and their connections are inactivated.}
	\label{fig:gnn_sampling}
\end{figure}
